---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: yes
---


## Loading and preprocessing the data

Loading the needed libaries
```{R, message=FALSE, warning=FALSE}
library("dplyr")
library("lubridate")

```

Reading the data and transforming data.frame to tibble for better performance and nice prints. 
Setting up the date correctly. 

```{R}
activity <- read.csv(unzip("activity.zip"))
activity <- tbl_df(activity)
activity <- mutate(activity,  date=ymd(date))
head(activity)
```

## What is mean total number of steps taken per day?

Calculating the total number of steps taken per day and plotting a histogram:
```{R}
activityByDate <- summarise(group_by(activity, date), sum(steps))
hist(activityByDate$`sum(steps)`, xlab = "steps", main ="Daily activity")
```

Calculating the mean and meadian for total number of steps taken per day and ignoring missing values. 

The mean number of steps per day: 
```{R}
meanSteps <- mean(activityByDate$`sum(steps)`, na.rm = TRUE)
meanSteps
```

The median number of step per day: 
```{R}
medianSteps <- median(activityByDate$`sum(steps)`, na.rm = TRUE)
medianSteps
```

## What is the average daily activity pattern?

Calculating the mean number of steps per interval 
```{r}
activityByInter <- summarise(group_by(activity, interval), steps = mean(steps, na.rm=TRUE))
```

Plotting the daily activity pattern. 
```{r}
plot(activityByInter, type="l", xlab ="index of 5-minute interval identifier", main = "Daily activity pattern")

```

Finding the 5-minute interval that on average contains the maximum number of steps
```{r} 
activityByInter$interval[which.max(activityByInter$steps)]
```

## Imputing missing values
Calculating the number of missing values in the data

```{r}
missing <- sum(is.na(activity$steps))
proportion <- mean(is.na(activity$steps))

```

The steps data has `r missing`  rows with missing values which is `r proportion`% of the data.

The missing values are replaced with the interval mean.
```{r}
activityNa <-  activity %>%
    group_by(interval) %>% 
    mutate(steps = ifelse(is.na(steps), mean(steps, na.rm=TRUE), steps))
```

Re-doing the analyses with missing values filled.

```{r}
activityByDate <- summarise(group_by(activityNa, date), sum(steps))
hist(activityByDate$`sum(steps)` , xlab = "steps", main ="Daily activity")
```

Calculate the median and mean
```{R}
meanStepsNa <- mean(activityByDate$`sum(steps)`, na.rm = TRUE)
medianStepsNa <- median(activityByDate$`sum(steps)`, na.rm = TRUE)
```

Make a nice table of the results to high light the differences. 
```{r}
meanMedian <- data.frame( NAs.filed = c(meanStepsNa, medianStepsNa), 
                         NAs.removed = c(meanSteps, medianSteps), row.names = c("mean",                             "median"))
meanMedian
```

Based on the calculated means and medians adding the missing data has very little effect to the results. This is mainly because there is such a low amount of missing values (`r missing`) compared to the total amount of values available (17 568). 

## Are there differences in activity patterns between weekdays and weekends?

```{r, warning=FALSE}
activityNa <-mutate(activityNa, weekend= ifelse(wday(date) == c(1,7), 
                                                "weekend", "weekday"))

```


Loading the needed libary
```{r, message=FALSE, warning=FALSE} 
library("lattice")
```

Plotting the activity and separating weekends and weekdays to separate plots. 
```{r}
activityWeek <- summarise(group_by(activityNa, weekend, interval), steps = mean(steps))

with(activityWeek, (xyplot(steps~interval|weekend, type="l", layout = c(1,2), xlab = "Interval", ylab = "Number of Steps")))

```


